---
title: "DSA 8020 Project 1"
author: "saransh rakshak"
output:
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libtoolkit}
library(dplyr)
library(ggplot2)
library(MASS)
library(mgcv)
library(car)
```

# Instructions

For each problem, your answer should include the following elements:

- A descriptive summary of the data using summary statistics and at least one plot/graph or table. Give your description in paragraph form.

- An appropriate inferential analysis to address the question. This may be a confidence interval, hypothesis test, or both. You can use the $\alpha$ value/confidence level of your choice. If you perform a hypothesis test, state the hypotheses, test statistic, p-value, decision, and conclusion. If you create a confidence interval, state the confidence level and provide an interpretation of the interval.

- A thoughtful conclusion that summarizes your findings. You should also discuss any limitations or complications of your analysis. This should include the following:

    * An assessment of whether the modeling assumptions are reasonable.
    
    * Discussion of any unusual features of the data. For example, do outliers/high leverage points affect your results?
    
---

# Problem 1: Gala

- Perform a detailed regression analysis using the gala data set. Use Endemics as the response variable, excluding Species from this analysis, to address the following questions:

**Code:**
```{r loadgaladata}
library(faraway)
data(gala)

# Excluding Species from analysis
gala <- subset(gala, select = -Species)

head(gala)
```

### Data Format

Source: M. P. Johnson and P. H. Raven (1973) "Species number and endemism: The Galapagos Archipelago revisited" Science, 179, 893-895

The dataset contains the following variables

Species (Excluded from analysis)
- the number of plant species found on the island

Endemics
- the number of endemic species

Area
- the area of the island (km$^2$)

Elevation
- the highest elevation of the island (m)

Nearest
- the distance from the nearest island (km)

Scruz
- the distance from Santa Cruz island (km)

Adjacent
- the area of the adjacent island (square km)

---

## 1. Provide a descriptive summary of the data set.

**Code:**
```{r}
summary(gala)
```

```{r}
# plot for endemics (response var)
# hist
ggplot(gala, aes(x = Endemics)) + 
  geom_histogram(binwidth = 10, color = "black", fill = "blue") +
  labs(title = "Distribution of Endemsics", x = "endemsics", y = "freq")

# density
ggplot(gala, aes(x = Endemics)) +
  geom_density(color = "red", fill = "red") +
  geom_rug() +
  labs(title = "Density of endemsics", x = "endemsics")

# plotting dist and density for Area
ggplot(gala, aes(x = Area)) +
  geom_histogram(binwidth = 100, color = "black", fill = "blue") +
  labs(title = "Dist. Area", x = "area", y = "freq")

ggplot(gala, aes(x = Area)) +
  geom_density(color = "red", fill = "red") +
  geom_rug() +
  labs(title = "Density Area", x = "area (km^2)")

# plots for Elevation
ggplot(gala, aes(x = Elevation)) +
  geom_histogram(binwidth = 50, color = "black", fill = "blue") +
  labs(title = "Dist. Elevation", x = "elevation", y = "freq")

ggplot(gala, aes(x = Elevation)) +
  geom_density(color = "red", fill = "red") + geom_rug() +
  labs(title = "Density of elevation", x = "elevation")

# Nearest
ggplot(gala, aes(x = Nearest)) +
  geom_histogram(binwidth = 0.5, color = "black", fill = "blue") +
  labs(title = "Dist. Nearest", x = "nearest", y = "freq")

ggplot(gala, aes(x = Nearest)) +
  geom_density(color = "red", fill = "red") +
  geom_rug() +
  labs(title = "Density of Nearest", x = "Nearest")

# Scruz
ggplot(gala, aes(x = Scruz)) +
  geom_histogram(binwidth = 30, color = "black", fill = "blue") +
  labs(title = "Dist. Scruz", x = "scruz", y = "freq")

ggplot(gala, aes(x = Scruz)) +
  geom_density(color = "red", fill = "red") +
  geom_rug() +
  labs(title = "Density of Scruz", x = "scruz")

# Adjacent
ggplot(gala, aes(x = Adjacent)) +
  geom_histogram(binwidth = 100, color = "black", fill = "blue") +
  labs(title = "Dist. Adjacent", x = "adjacent", y = "freq")

ggplot(gala, aes(x = Adjacent)) +
  geom_density(color = "red", fill = "red") +
  geom_rug() +
  labs(title = "Density of Adjacent", x = "Adj.")

```


**Answer:**

- (Endemics): Our response variable, Endemics, shows high variability with a large range (0 to 95). This indicates either that there is either a diverse and varying degrees of endemics in species observations. Our plot shows Endemics to appear right-skewed, showing that some islands have significantly more endemic species than others.

- (Area, Adjacent): Both variables show significant variation, with a high max compared to mean values, and show right-skewed distributions. This suggest potential outliers confirmed on the distribution graphs. 

- (Scruz): Also shows moderately high variation, outliers can be seen in distribution graph however the presence of multiple values supports theory that these outliers are in fact significant values, as opposed to what we saw with Area & Adjacent.

- (Elevation, Nearest): Both show more balanced distributions compared to our other variables, but still their values lie in a wide range.

---

## 2. Is at least one of the predictors useful in predicting the response (Endemics)?

Establishing an alpha ($\alpha$) value of 0.05. 

**Code:**
```{r}
# getting p-vals
model <- lm(Endemics ~ Area + Elevation + Nearest + Scruz + Adjacent, data = gala)
summary(model)
# corelation
cor(gala[, c("Area", "Elevation", "Nearest", "Scruz", "Adjacent")], gala["Endemics"])

# Checking model assumptions
# residual vs fitted
plot(model, which = 1)

# qq plot
qqnorm(resid(model))
qqline(residuals(model))
```

**Answer:**

Allowing an $\alpha = 0.05$, we can see that...

- Elevation (p-val: 5.83e-08) and Adjacent (p-val: 5.50e-05) are useful in predicting the response (Endemics).

- Area (p-val: 0.0732) is slightly above our alpha 0.05, suggesting that Area has some predictive value but is not statistically significant.

- Scruz (p-val: 0.2044) and Nearest (p-val: 0.9066) have little to no meaningful relationship with Endemics that can be leveraged in prediction. 


---

## 3. Perform model selection to determine whether all the predictors help to explain the response.

**Code:**
```{r}
# remove pointless vars iteratively based on AIC
stepmodel <- stepAIC(model, direction = "both")
summary(stepmodel)
```

**Answer:**

- Using step-wise AIC selection, we can get a final optimal model composed of variables...

    - Adjacent: (p-val: 2.01e-05) shows to be highly significant in explaining the number of endemic species.
    
    - Elevation: (p-val: 1.73e-08) shows to be highly significant in explaining the number of endemic species.
    
    - Area: (p-val: 0.0591) is borderline insignificant at our alpha = 0.05 but can still be slightly useful in a less strict model.
    
    - Scruz: (p-val: 0.1201) is not statistically significant at our alpha = 0.05, and will not help much in predicting the response.
    
- Our step-wise AIC removed variable Nearest as it was shown to increase AIC, reducing response prediction in our final model.

In summation, all variables were not useful in predicting the response, and variable Nearest was removed using step-wise AIC, giving a new optimized model with:
    
> Endemics ~ Area + Elevation + Scruz + Adjacent
    
---

## 4. Perform a Box-Cox transformation to determine if there is a need to transform the response.

**Code:**
```{r}
# to avoid 0.00value stopping boxcox(), shifting by insignificant amount
gala$Endemics <- gala$Endemics + 0.000001

# boxcox of stepwise AIC optimized model
bc <- boxcox(stepmodel)
plot(bc)
```

- As we get an approximate $\lambda$ value of near 0.5, I will apply a square root transformation to our response variable Endemics.

```{r}
bc_sqrt_endemics <- sqrt(gala$Endemics)
bc_sqrt_model <- lm(bc_sqrt_endemics ~ Area + Elevation + Scruz + Adjacent, data = gala)
summary(bc_sqrt_model)

# residual vs fitted
plot(bc_sqrt_model, which = 1)

# normal QQ
qqnorm(resid(bc_sqrt_model))
qqline(residuals(bc_sqrt_model))
```

- Both plots show significant improvements over our original model. There is significantly less spread in Residuals as well as a more linear relationship in our QQ plot. 

- Residual plots for both models(original and transformed) showed no significant patterns or heteroscedasticity; transformation doesn't violate assumptions of linear  regression.


---

## 5. Compare a generalized additive model fit with the linear regression fit and explain why you preferred one model over the other.

**Code:**
```{r}
gam_model <- gam(bc_sqrt_endemics ~ s(Area) + s(Elevation) + s(Scruz) + s(Adjacent), data = gala)

AIC(bc_sqrt_model, gam_model)
BIC(bc_sqrt_model, gam_model)
```

**Answer:**

- Our GAM model shows considerable improvement in degrees of freedom, (14 vs 6) thus proves to be more flexible.

- Our GAM model also shows considerable improvement in AIC (97.4 vs 109.7 for Box-Cox transformation only) and slight improvement in BIC (117.1 vs 118.2 for Box-Cox transformation only).

- Thus, our GAM model will be preferred moving forward.

---

## 6. How well does the chosen model fit the data? Quantify the fit numerically. Additionally, check the model assumptions and summarize your findings.

**Code:**
```{r}
# comparing with Box-Cox sqrt model
lm_dev <- deviance(bc_sqrt_model)
null_lm_dev <- deviance(update(bc_sqrt_model, ~ -1))
dev_explained_lm <- 1 - (lm_dev / null_lm_dev)
dev_explained_lm

# gam model
gam_dev <- deviance(gam_model)
null_gam_model <- gam(bc_sqrt_species ~ 1, data = gala)
null_gam_dev <- deviance(null_gam_model)
dev_explained_gam <- 1 - (gam_dev / null_gam_dev)
dev_explained_gam
```

- There is a considerable improvement in deviance explained by the GAM model, 0.97 vs 0.94 with our Box-Cox transformation only.

- This confirms the GAM model to better fit the data and be the optimal choice.

```{r}
# residual vs fitted
gam.check(gam_model)

# normal QQ
qqnorm(residuals(gam_model))
qqline(residuals(gam_model))
```

- in



```{r}
summary(gam_model)
```



- Our QQ plots show our residuals to be normally distributed, showing no violation in model assumptions.

---

## 7. Predict the response value when all the chosen predictors take their median value. Quantify the prediction uncertainty.

**Code:**

---

## 8. Provide a thoughtful conclusion that summarize your findings.

**Code:**

---


# Problem 2: Salaries

- Perform a detailed analysis using the Salaries data set to comment on the salary differences between male and female faculty members.

**Code:**
```{r loadsaldata}
library(carData)
data(Salaries)
```

### Data Format

References: Fox J. and Weisberg, S. (2019) An R Companion to Applied Regression, Third Edition, Sage.

A data frame with 397 observations on the following 6 variables.

rank
- a factor with levels AssocProf AsstProf Prof

discipline
- a factor with levels A (“theoretical” departments) or B (“applied” departments).

yrs.since.phd
- years since PhD.

yrs.service
- years of service.

sex
- a factor with levels Female Male

salary
- nine-month salary, in dollars.

---

## 1. Provide a descriptive summary of the data set.

**Code:**
```{r}
head(Salaries)
summary(Salaries)
```

**Answer:**

- The majority of the staff roster is made up of Full Professors (266).

- There are slightly more Applied discipline (216) than Theoretical (181).

- Mean years since Phd is 22.31 years, while mean years of service is 17.61 years.

- Male staff makes up an overwhelming majority of the staff roster.

- Salary varies widely, with a median of 107k, and a maximum of 231k.

    - Mean salary is slightly higher at 113k --> potential positive skewing salary distribution.

- Distribution indicates some professors earn significantly more, most likely based on rank and years of service.

---

## 2. Comment on whether there is a need to include interactions between these categorical predictors. Select your model and justify your choice.

**Code:**
```{r}
ggplot(Salaries, aes(x = salary, fill = rank)) +
  geom_density(alpha = 0.3) + 
  facet_wrap(~rank) + 
  labs(title = "Salary dist by Rank", x = "salary", y = "density")
```

```{r}
# salary distribution by discipline and rank
ggplot(Salaries, aes(x = salary, fill = interaction(rank, discipline))) +
  geom_density(alpha = 0.3) + 
  facet_wrap(~discipline) + 
  labs(title = "Salary dist Discipline and Rank", x = "salary", y = "density")
```

```{r}
# salary distribution years since pdf (cat)
ggplot(Salaries, aes(x = factor(yrs.since.phd), y = salary)) +
  geom_boxplot() +
  labs(title = "Salary dist by yrs.since.phd", x = "years since phd", y = "salary")
```

```{r}
# salary dist by sex
ggplot(Salaries, aes(x = salary, fill = sex)) +
  geom_density(alpha = 0.3) +
  labs(title = "salary dist by sex", x = "salary", y = "density")
```


---

## 3. Perform a log transformation on response to determine if there is any improvement in terms of model fit.

**Code:**


---

## 4. How well does the chosen model fit the data? Quantify the fit numerically. Additionally, check the model assumptions and summarize your findings.

**Code:**

---

## 5. Write a thoughtful conclusion to comment on the salary differences between male and female faculty.

**Code:**





---









