---
title: "DSA 8020 R Session 8: Replication, Blocking, and Randomization"
author: "Whitney"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    fig_width: 5
    fig_height: 4
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objective
This \texttt{R} session demonstrates the principles of *replication*, *blocking*, and *randomization* in the design and analysis of experiments. By the end of this exercise, you should understand:

+ The role of *randomization* in ensuring unbiased results.
+ Why *replication* is crucial for estimating variability.
+ How *blocking* helps reduce confounding variables.


## Experimental Scenario

In this experiment, we will use **paper helicopters** to study the effect of two **helicopter wing lengths** on **flight time (seconds)**. The experiments are performed by **different experimenters**, which introduces a potential blocking factor.


## Load Necessary Libraries

```{r,message=FALSE}
library(dplyr)
library(ggplot2)
```

## Step 1: Experimental Design

### 1.1 Define Factors and Levels

+ *Treatment:* Wing length ("Short", 5 cm or "Long", 10 cm)
+ *Blocking Factor:* Experimenter (1 or 2)
+ *Response Variable:* Flight time (seconds)

Each treatment level will be **replicated** 10 times. Moreover, we will **block** the experiment by assigning a single experimenter to test all treatment levels within a block to control for variability due to individual differences in timing.

### 1.2 Randomization

To minimize bias, we will **randomize** the order of helicopter drops.

## Step 2: Data Collection

### 2.1 Experiment configuration

```{r}
rep <- 10
treatment <- rep(c("Short", "Long"), each = rep)
block <- rep(c("Experimenter 1", "Experimenter 2"), times = rep)
cbind(treatment, block)
```

### 2.2 Simulating Flight Time Data

For demonstration, we simulate flight times assuming variability:

+ Short wings: $\mu$ = 2.6 sec, $\text{sd} = \sqrt{\text{Var}}$ = 0.5 sec
+ Long wings: $\mu$ = 3.2 sec, $\text{sd} = \sqrt{\text{Var}}$ = 0.6 sec

Additionally, we assume there is a measurement error for experimenters, with the first experiment tending to underestimate ($\mu$ = -0.3 sec) while the second overestimates ($\mu$ = 0.3 sec).
Finally, we assume there is an order effect that negatively biases the flight time over runs (and therefore, it is important to randomize the runs to mitigate such a bias).


```{r}
set.seed(123)  # For reproducibility
n <- length(treatment)
effect <- c(rnorm(rep, mean = 2.6, sd = 0.5),  # Short
            rnorm(rep, mean = 3.2, sd = 0.6))  # Long
experimenter <- rnorm(n, mean = rep(c(-0.3, 0.3), rep), sd = 0.1)
order <- seq(0.5, -0.5, len = n)
time1 <- effect + experimenter + order 
# Randomize the order
run_order <- sample(n)
time2 <- effect + experimenter + order[run_order] 

data1 <- data.frame(Treatment = treatment, Block = block, Response = time1)
data2 <- data.frame(Treatment = treatment, Block = block, Response = time2)
```

## Step 3: Summarizing and Visualizing the Data

Without and with randomization

```{r}
tapply(data1$Response, list(data1$Treatment, data1$Block), mean)
tapply(data2$Response, list(data1$Treatment, data1$Block), mean)
```
Without randomization

```{r}
ggplot(data1, aes(x = Treatment, y = Response, color = Block)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.7) +
  theme_minimal() +
  labs(title = "Effect of Wing length on Flight time",
       x = "Wing length",
       y = "Flight time (seconds)")
```

With randomization

```{r}
ggplot(data2, aes(x = Treatment, y = Response, color = Block)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.7) +
  theme_minimal() +
  labs(title = "Effect of Wing length on Flight time",
       x = "Wing length",
       y = "Flight time (seconds)")
```


## Step 4: Analyze the Data

### 4.1 Perform ANOVA to Assess Treatment and Block Effects

Model 1: did not account for the block effect and was not randomized to eliminate the run order bias

```{r}
model1 <- aov(Response ~ Treatment, data = data1)
summary(model1)
coef(model1)
```

Model 2: did not account for the block effect but was randomized to eliminate run order bias

```{r}
model2 <- aov(Response ~ Treatment, data = data2)
summary(model2)
coef(model2)
```

Model 3: did account for the block effect but was not randomized to eliminate the run order bias

```{r}
model3 <- aov(Response ~ Treatment + Block, data = data1)
summary(model3)
coef(model3)
```
Model 4: did account for the block effect and was randomized to eliminate the run order bias

```{r}
model4 <- aov(Response ~ Treatment + Block, data = data2)
summary(model4)
coef(model4)
```

### 4.2 Interpretation

+ *Replication* ensures that variability can be measured within treatments.
+ *Blocking* accounts for the effect of different experimenters.
+ *Randomization* prevents systematic biases in treatment allocation.


## Conclusion

By conducting this experiment with proper replication, blocking, and randomization, we ensure a robust and reliable analysis of treatment effects. The ANOVA results indicate whether wing length significantly affects flight time while controlling for experimenter variation.


