---
title: "Project 3 - DSA 8020"
author: "Saransh Rakshak"
date: "Due: Thursday May 1, 2025"
output:
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r loadlibs, include= FALSE}
library(fields)
library(tsfeatures)
```


# Project Description

The ERA-Interim is a global atmospheric reanalysis dataset. Reanalysis refers to the process of producing spatially and temporally gridded datasets through data assimilation, primarily for climate monitoring and analysis. In the R data file **NA_MaxT_ERA.RData**, you will find daily maximum temperature values from *1979* to *2017*, with a spatial resolution of approximately 80 km. The original global dataset has been subsetted to a region covering the contiguous United States and extending into Canada and Mexico.

In Problems 1 and 2, you are required to analyze an ERA-Interim average temperature time series at yearly and monthly temporal resolutions at a spatial location of your choice.

---

# Problem 1. Analyzing Yearly Average Temperatures Time-Series

### 1. The following steps are needed to extract the annual mean temperatures at the specified location from the original dataset with daily temperatures at thousands of locations:

#### a. First, use the following R script to load the R data object:

```{r}
load("NA_MaxT_ERA.RData")
NA_MaxT <- NA_MaxT - 273.15 # change from Kelvin to Celsius
nlon <- length(NA_lon); nlat <- length(NA_lat)
nmon <- 12; nyr <- 39 # numbers of month/year
```

#### b. Second, choose a location of interest (with the information of longitude and latitude) and identify the nearest ERA grid cell to the chosen location. Below you can find a script to get the correct ERA grid cell for Clemson:

```{r}
# Clemson Example
Clemson.lon.lat <- c(-82.8374, 34.6834) # (lon,lat)
dist2Clemson <- rdist.earth(
  matrix(Clemson.lon.lat, 1, 2), 
  expand.grid(NA_lon, NA_lat), 
  miles = F
  )
id <- which.min(dist2Clemson)
(lon_id <- id %% nlon)
(lat_id <- id %/% nlon + 1)

# Check
(rdist.earth(matrix(Clemson.lon.lat, 1, 2),
matrix(c(NA_lon[lon_id], NA_lat[lat_id]), 1, 2), miles = F)
== min(dist2Clemson))
```

```{r}
Monterey.lon.lat <- c(-121.8979, 36.5973) # (lon,lat)
dist2Monterey <- rdist.earth(
  matrix(Monterey.lon.lat, 1, 2), 
  expand.grid(NA_lon, NA_lat), 
  miles = F
  )

id <- which.min(dist2Monterey)

(lon_id <- id %% nlon)
(lat_id <- id %/% nlon + 1)

(rdist.earth(
  matrix(Monterey.lon.lat, 1, 2),
  matrix( c(NA_lon[lon_id], NA_lat[lat_id]), 1, 2),
  miles = F
  )== min(dist2Monterey))
```

#### c. Third, aggregate the daily data to the annual average time series. You could use the following script to achieve this:

```{r}
time_series_daily <- NA_MaxT[lon_id, lat_id,]
print("Daily Time Series summary: ")
print(summary(time_series_daily))
```

```{r}
time_series_yearly <- tapply(time_series_daily, list(yr), mean)
print("Yearly Time Series summary: ")
print(summary(time_series_yearly))
```

### 2. Describe the main features of the annual time series data you extracted.

```{r}
# using package tsfeatures for feature autoextraction
yearly_features <- tsfeatures(time_series_yearly)
print(yearly_features)
```

Our time series feature analysis using *tsfeatures* for the annual series reveals the following:

| Feature | Value | Analysis |
|---------|-------|----------|
|trend    |0.4337 |Moderately strong trend, indicates presence of long term pattern |
|spike    |0.00042|Extremely low, indicates no significant anomalies or jumps in any given year|
|linearity|0.08572|Moderately low - some possibility model is linear|
|curvature|2.27771|High value strongly indicates non-linearity|
|entropy  |0.9574 |High value, indicates limited predictability and complex relations between the data|
|x_acf1   |0.4117 |Moderate indication of autocorrelation, any given years values show to depend on its previous year|

### 3. Is it reasonable to assume that there is a linear trend? If so, estimate and remove the trend to get the detrended time series.

```{r}
plot(time_series_yearly, main = "Annual Time Series Plot", ylab = "Observed Value", xlab = "Time")
```


Our graph does not reveal much significant information. However, based on our previous results from *tsfeatures*, it is still somewhat reasonable to expect the possibility of a linear trend:

    1. trend: value 0.43 shows significant enough trend in data to expect some prediction.
    
    2. linearity: while this is not 0, it is still low enough to keep linearity as a reasonable possibility.
    
    3. curvature: even though value is high (indicating a non-perfect linear fit), it is still significant enough to keep linearity a potential assumption, especially early on in the analysis.
    
Since a linear trend is not out of question for our time series, we will now estimate its value.

```{r}
yrs <- unique(yr)

# fitting linear model
lm_annual <- lm(time_series_yearly ~ yrs)
summary(lm_annual)
```

```{r}
plot(
  x= yrs,
  y= time_series_yearly,
  type= "o", # show actual measured points with their continuity over time 
  main = "Yearly Time Series(Trend)",
  ylab = "Observed Value",
  xlab = "Time",
  panel.first= grid()
  )
abline(lm_annual, col = "red", lwd = 2)
```

Our linear model fit shows a p-value of 0.64, showing that a linear trend is statistically insignificant - leading us to fail to reject the null hypothesis and conclude there is no strong proof of linear behavior in the data. This is further confirmed by our Yearly Time Series(Trend) plot, which shows the data fluctuating significantly around our linear trend line (red) as well as many clearly non-linear patterns. It may be possible that the data could have a different measurable trend such as cyclical behavior, as is not unusual with environmental data.


<!-- TODO: Estimate and detrend -->












