---
title: 'DSA 8020 R Session 0: A Quick Introduction to R'
author: "Whitney Huang"
date: ""
output:
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This \texttt{R} session is modified from a R tutorial given by Dr. Doug Nychka at Colorado School of Mines [[link]](https://dnychka.github.io/).


## Basic

### What is in the workspace?


```{r}
ls()
```

### Where is my \texttt{R} working directory?

```{r}
getwd()
```


### What is in my working directory?

```{r}
dir()
```

### "$c$" combines sets of numbers (or datasets)

```{r}
x <- c(2, 3, 20)
```

*Note*: \texttt{R} is case sensitive. Type "X" in \texttt{R} console and then click
 Enter to see what happen

### Now recheck workspace

```{r}
ls()
# print out x
x
```

### Reassign "x" to another name

```{r}
x2 <- x
ls()
```

### Remove "x" and create another object "x3"

```{r}
rm(x)
x3 <- c(3, 4, 5)
```


### Question: how would you combine "x2" and "x3" to make a new data set ?

```{r}
#give a try here
```

## Arithemtic

### Add numbers in \texttt{R}

```{r}
A <- 2
B <- 10
Y <- A + B

A <- c(2, 3, 4)
B <- c(10, 100, 1000)
Y <- A + B
Y # note that the numbers have been added row by row like a spread sheet.
```


### Some other operations

```{r}
2^4
2 * (1 + 4)
sqrt(81)
exp(2)
```

### Generating a sequence 

```{r}
1:10
-5:5
```

*Question*: How would you generate the values in order 5 to 1?

### Generating sin wave

```{r}
x <- 0:100
# hundred values between 0 and 2*pi
x <-  2 * pi * (x / 100)
y <- sin(x)
# plot the sin wave 
plot(x, y)
# change the plot to connect points with a line instead of points
plot(x, y, type = "l")
# another way of creating the x
x <- seq(0, 2 * pi, length.out = 101)
```


### Use \texttt{R} to generate random values 

```{r}
# generating 10 random numbers between 0 and 1 
U <- runif(10)
U
# generating 30 random numbers from a standard normal distribution
V <- rnorm(n = 30, mean = 0, sd = 1)
hist(V)
qqnorm(V); qqline(V)
```

## Subsetting

### Load a data set 

```{r}
load("/home/sar-home/Clemson/DSA-8020/datasets/BT.RData")
# copy for easy typing
BT <- BoulderJuneTemperature$Temp
BAll <- BoulderJuneTemperature
head(BT); head(BAll, 10)
```


### Print the first 10 values
```{r}
BT[1:10]
```

### An indicator for all values over 70
```{r}
ind70 <- BT > 70
ind70
```

 
```{r}  
# temperatures with values over 70
BT[ind70]
```

```{r}  
# the years with values over 70
BAll$Year[ind70]
```

*Question*: How many years exceed 70 degrees?

### Working with these data as a matrix

```{r}
dim(BAll)
# This is the first row and first column
BAll[1, 1]
# first row
BAll[1,]
# first column
BAll[, 1]
# column with Year (this is also column 1)
BAll[, "Year"]
# second column (could also use "Temp" to refer to this)
BAll[, 2]
# rows 10 through 20
BAll[10:20,]
```

*Exercise*: Plot the temperatures by year
  
## Apply Functions in \texttt{R}

### *apply* functions

1. a family of functions in R which allow you to repetitively perform an action on multiple chunks of data

2. run faster than loops and often require less code. 

Let's take a look at some examples

### Load the Boulder temperature data set into R 
```{r}
load("/home/sar-home/Clemson/DSA-8020/datasets/BoulderTemperature.RData") # monthly mean temperatures
dim(BoulderTemperature)
# check out first row
BoulderTemperature[1,]
#extract 1991 - 2010
yr <- rownames(BoulderTemperature)
index <- which(yr %in% 1991:2010) 
tempData <- BoulderTemperature[index,]
# check this out 
tempData
```

### The "*apply*" function

```{r}
# means  by rows of this table
byYear <- apply(tempData, 1, FUN = mean) # by rows, 1 = first index
byYear
rowMeans(tempData)
# means by columns
byMonth <- apply(tempData, 2, FUN = mean) # by cols, 2 = second index
byMonth
colMeans(tempData)
```


## Writing Functions in \texttt{R}


### Finding the inter quartile range (IQR)

```{r}
# 75% quantile  
BT75 <- quantile(BT, .75)
#Question: Find the interquartile range 75% - 25% quantiles
# and check this against the built in function
IQR(BT)
```

### Building your own function

Here is a function that adds the squares of two numbers. It has three parts, the *calling arguments*, the *body* where you do the work and then *returning any results*.

```{r}
myFun <- function(a, b){
    result <- a^2 + b^2
    return(result)
}

test1 <- myFun(2, 3)
test1
test2 <- myFun(1:5, 11:15)
test2
```

Note that the "a", "b" and result are only used inside the function and
do not appear in your workspace.
Also since the body is normal R code, this works for vectors automatically.

### Building your own IQR function

```{r}
myIQR <- function(y){
  IQR <- quantile(y, .75, names = FALSE) - quantile(y, .25, names = FALSE)
  return(IQR)
}
myIQR(BT)
```

### Modify this function to work with NAs
```{r}
test <- c(BT, NA)  
myIQR <- function(y, na.rm = FALSE){
            IQR <- quantile(y, .75, names = FALSE, na.rm = na.rm) - 
                   quantile(y, .25, names = FALSE, na.rm = na.rm) 
            return(IQR)
}
myIQR(test, na.rm = T)
```

### Adding warning message 

```{r}
myIQR <- function(y, na.rm = FALSE){
# an example of adding a warning     
    if(na.rm){
        warning("NAs removed from the data")
    }
            IQR <- quantile(y, .75, names = FALSE, na.rm = na.rm) - 
                   quantile(y, .25, names = FALSE, na.rm = na.rm) 
            return(IQR)
} 
myIQR(test, na.rm = T)
```

