---
title: "DSA 8020 R Session 4: Multiple Linear Regression III"
author: "Whitney"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    fig_width: 7
    fig_height: 6.5
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(faraway)
data(gala)
galaNew <- gala[, -2]
```

## Model Selection

### All Subset Selection

```{r, message=FALSE}
library(leaps)
models <- regsubsets(Species ~ ., data = galaNew)
summary(models)
```

### Reporting model selection criteria

```{r}
res.sum <- summary(models)
criteria <- data.frame(Adj.R2 = res.sum$adjr2,
  Cp = res.sum$cp, BIC = res.sum$bic)

criteria

plot(2:6, criteria$Cp, las = 1, xlab = "p", ylab = "Cp",
     pch = 16, col = "gray", ylim = c(1, max(criteria$Cp)))
abline(0, 1)

plot(2:6, criteria$Adj.R2, las = 1, xlab = "p", ylab = "", pch = 16, col = "gray",
     main = expression(R['adj']^2))
points(5, criteria$Adj.R2[4], col = "blue", pch = 16)

plot(2:6, criteria$BIC, las = 1, xlab = "p", ylab = "", pch = 16, col = "gray", main = "BIC")
points(3, criteria$BIC[2], col = "blue", pch = 16)
```

### Backward Selection

Starts with all the predictors and
then removes predictors one by one using some criterion


```{r}
full <- lm(Species ~ ., data = galaNew)
step(full, direction = "backward")
```

### Stepwise Selection

A combination of backward elimination and forward selection can involve adding or deleting predictors at each stage


```{r}
step(full, direction = "both")
```

## Model Diagnostics

### Residual Plot

```{r}
mod <- lm(Species ~ Elevation + Adjacent, data = galaNew)
plot(mod$fitted.values, mod$residuals, pch = 16, col = "blue")
abline(h = 0, col = "red")
```

### Residual Histogram/QQplot

These are used for assessing normality of residuals

```{r}
par(las = 1)
hist(mod$residuals, 5, prob = T, col = "lightblue", border = "gray")
xg <- seq(-200, 200, 1)
sd <- sd(mod$residuals)
yg <- dnorm(xg, 0, sd)
lines(xg, yg)

plot(qnorm(1:30 / 31, 0, sd), sort(mod$residuals), pch = 16,
     col = "gray", xlab = "Normal Quantiles", ylab = "Residuals")
abline(0, 1)
```

### Leverage

Detecting *extreme* predictor values

```{r}
step_gala <- step(full, trace = F)
X <- model.matrix(step_gala)
H <- X %*% solve((t(X) %*% X)) %*% t(X)
diag(H)
lev <- hat(X)
hatvalues(step_gala)

high_lev <- which(lev >= 2 * 3 / 30)
attach(gala)
par(las = 1)
plot(Elevation, Adjacent, cex = sqrt(5 * lev), col = "blue", ylim = c(0, 5000))
points(Elevation[high_lev], Adjacent[high_lev], col = "red", pch = 16,
       cex = sqrt(5 *lev[high_lev]))
```

### Standardized Residuals

```{r}
gs <- summary(step_gala)
gs$sig
studRes <- gs$res / (gs$sig * sqrt(1 - lev))

rstandard(step_gala)

par(las = 1)
plot(step_gala$fitted.values, studRes, pch = 16, col = "blue",
     ylab = expression(r[i]), main = "Studentized Residuals", xlab = "")
abline(h = 0, lty = 2, col = "gray")
```


### Studentized (Jackknife) Residuals

```{r}
jack <- rstudent(step_gala)

par(las = 1)
plot(jack, pch = 16, cex = 0.8, col = "blue", main =" Jacknife Residuals ",
     xlab = "", ylab = "")
abline(h = 0, lty = 2, col = "gray")
```


### Identifying Influential Observations: DFFITS

```{r, message=F}
dffits(step_gala)
library(olsrr)
ols_plot_dffits(step_gala)
```

### Identifying Influential Observations: Cook’s Distance

```{r}
cooks.distance(step_gala)

par(mfrow = c(2, 1), mar = c(3.8, 3.8, 1.2, 0.5), mgp = c(2.5, 1, 0), las = 1)
plot(step_gala, which = 4:5)
```

### Response transformation

```{r}
par(las = 1)
plot(step_gala$fitted.values, step_gala$residuals, 
     pch = 16, cex = 0.8, col = "blue", main =" Residuals ",
     xlab = expression(hat(Y)), ylab = expression(e))
abline(h = 0, lty = 2, col = "gray")

sqrt_fit <- lm(sqrt(Species) ~ Elevation + Adjacent)

plot(sqrt_fit$fitted.values, sqrt_fit$residuals, 
     pch = 16, cex = 0.8, col = "blue", main =" Residuals ",
     xlab = expression(hat(Y)), ylab = expression(e))
abline(h = 0, lty = 2, col = "gray")
```

### Box-Cox Transformation

```{r,message=FALSE}
library(MASS)
par(las = 1)
boxcox <- boxcox(step_gala, plotit = T, lambda = seq(-0.25, 0.75, by = 0.05))
```

