---
title: 'DSA 8020 R Lab 10: Interpolation of Spatial Data'
author: "Saransh Rakshak"
date: "Due: Thurs. April 24"
output:
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(gstat)
library(sp)
```

---

## 1. Load the data into R. How many observations are in this spatial data set?

**Code:**

```{r}
load("data/RLab10.RData")
length(s)
```

**Answer:**

There are 400 observations in spatial data set 's'.  
 
---

## 2. Plot the spatial data.

**Code:**

```{r, fig.asp=1}
ggplot(s, aes(x= x, y= y)) +
  geom_point(color = "blue") +
  labs(title= "Spatial Data",
       x= "X",
       y= "Y"
       ) +
  theme_minimal()
```


---

## 3. Plot the spatial variogram and briefly describe your findings.

**Code:**

```{r}
df <- as.data.frame(s)
colnames(df) <- c("x", "y")
df$z <- sin(df$x + df$y)
coordinates(df) <- ~ x + y

# variogram, mean func constant mean=1
v.emp <- variogram(z ~ 1, data= df)
plot(v.emp, main= "Spatial Variogram (Constant Mean)")
```

**Answer:**

- Shows increasing semivarience as distance increases, showing positive spatial autocorrelation. 

- Little to no nuggeting, shows low noise for measurements. 

- No still of curve, shows data does not yet have a spatial dependence. Will need to extend distances to greater than currently sampled to see signs of spatial dependence.


---

## 4. Fit a Spatial Gaussian process model to the data. Describe the models for the mean function and the covariance function, and report the estimated parameters.

**Code:**

*Mean Function Model*

```{r}
# mean function model - linear trend surface model: z~x+y
v.emp <- variogram(z ~ x + y, data = df)
plot(v.emp, main = "Empirical Variogram w/ Trend (Mean Model: z~x+y)")
```

*Covariance Function Model*

```{r}
# fit Gaus to empirical variogram
v.fit <- fit.variogram(v.emp, model = vgm(psill = 0.03, 
                                          model = "Gau", 
                                          range = 0.2, 
                                          nugget = 0) )

plot(v.emp, model = v.fit, main = "Fitted Gaussian Variogram (Covariance Model)")
```

```{r}
# estimated covariance vals
print(v.fit)
```


**Answer:**

*Mean Function Model*

> Model: $\mu (s) = \beta_0 + \beta_1 x + \beta_2 y$

> Code: variogram(z ~ x + y, data = df)

- $\mu (s)$ is our expected mean value of spatial process at distance (s).
    
- $\beta_0$ baseline when x and y equal 0.
    
- $\beta_1$ effect of x on mean.
    
- $\beta_2$ effect of y on mean.

*Covariance Function Model*

> Model: $Cov(h) = \sigma^2 exp(-(h / \theta)^2)$

> Code: fit.variogram(v.emp, model = vgm(psill = 0.03, model = "Gau", range = 0.2, nugget = 0))

- $\sigma_2$: partial sill.
    
- $\theta$: range.
    
- nugget: variance at distance 0.

*Estimated Parameters*

- Nugget: 5.482179e-06 very small, negligible.

- Partial Sill ($\sigma^2$): 7.059960e-03, is our variance of the residual spatial process after accounting for the z~x+y large scale trend. Value is extremely small, expected since trend surface already explains most of variation.

- Range ($\theta$): 0.4140399, shows points smaller than 0.414 distance units are strongly correlated. Points with greater distance units apart have rapidly declining correlation.
    

**Overall Conclusion**

Smooth decay in Gaussian model shows strong local spatial continuity. Our trend model *z ~ z + y* captured most of spatial variation, explains why partial sill and nugget have such small values. Gaussian covariance function continues to explain local correlation in residuals (within a 0.414 distance apart). Overall shows good specification in spatial model - our trend explains the big picture while covariance shows local detail.


---

## 5. Produce the spatial interpolated map with its associated estimation uncertainty.

**Code:**

```{r, fig.asp=1}
# grid to interpolate over
x.range <- range(df@coords[,1])
y.range <- range(df@coords[,2])

v.grd <- expand.grid(
  x = seq(x.range[1], x.range[2], length.out = 100),
  y = seq(y.range[1], y.range[2], length.out = 100)
)

coordinates(v.grd) <- ~x+y
gridded(v.grd) <- TRUE  # spatial grid

# gstat obj - kriging vpred
v.g <- gstat(id = "z", formula = z ~ 1, data = df, model = v.fit)
v.kriged <- predict(v.g, newdata = v.grd)

# interpolated
spplot(
  v.kriged["z.pred"],
  main = "Spatial surface map - interpolated",
  scales = list(draw = TRUE)
  )

# uncertainty (var) estimate
spplot(
  v.kriged["z.var"],
  main = "Estimation Uncertainty",
  scales = list(draw = TRUE)
  )
```





