---
title: "DSA 8020 R Lab 5: Analysis of covariance and Non-linear Regression"
author: "saransh rakshak"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Analysis of covariance: Salaries for Professors

The 2008-09 nine-month academic salary for Assistant Professors, Associate Professors, and Professors in a college in the U.S. was collected as part of the ongoing effort of the college's administration to monitor salary differences between male and female faculty members.

### Load the dataset

**Code:**

```{r}
library(carData)
library(dplyr)
data(Salaries)
head(Salaries)
```

**Description of the variables**

* \texttt{rank}: a factor with levels Assistant Professor ("AsstProf"); Associate Professor ("AssocProf"); Full Professor ("Prof")

* \texttt{discipline}: a factor with levels A ("theoretical" departments) or B ("applied" departments)

* \texttt{yrs.since.phd}: years since her/his PhD

* \texttt{sex}: a factor with levels "Female" and "Male"

* \texttt{salary}: nine-month salary, in dollars

### Exploratory Data Analysis

1. Identify the numerical variables and categorical variables in this data set

**Answer:**

Numerical Vars

> yrs.since.phd, salary

Categorical Vars

> rank, discipline, sex

---

2. Summarize each variable numerically and graphically, briefly describe your findings

**Code:**

```{r}
vars <- c("rank", "discipline", "yrs.since.phd", "sex", "salary")
data <- Salaries[, vars]
data <- data %>%
  mutate(rank = as.numeric(rank),
         discipline = as.numeric(discipline),
         sex = as.numeric(sex))

# numerical summary
summary(data)

# graphing summary
#par(mfrow=c(2, 3))
barplot(table(data$rank), main = "Barplot of Rank", xlab = "Rank", ylab = "Freq",
        names.arg = c("AssistantP", "AssociateP", "Professor"), 
        col = c("yellow", "orange", "red"))
barplot(table(data$discipline), main = "Barplot of Disciplines", xlab = "Discipline",
        ylab = "Freq", names.arg = c("A", "B"))
hist(data$yrs.since.phd, main = "Histogram of Years Since PHD", xlab = "Yrs")
barplot(table(data$sex), main = "Stacked Barplot of Sex", xlab = "Sex", 
        ylab = "Freq", names.arg = c("Female", "Male"), col = c("pink", "lightblue"))
boxplot(data$salary, main = "Boxplot of Salary", ylab = "Salary ($)")
```


**Answer:**

Rank

> There are more full professors than assistant or associate professors. 

Discipline

> Both A and B seem mostly balanced, with B having slightly greater occurances.

yrs.since.phd

> Most staff have between 10 and 40 years of elapsed time since their PhD.
> Distribution has slight right skew, with very few staff having 50 or more years since their PhD.

Sex

> Highly imbalanced, with Males greatly dominating the staff roster.

Salary

> Salaries range from $57k to $231k, with a median of $107k and an average salary of $113k.

---

3. Create a scatterplot matrix and briefly describe your findings

**Code:**

```{r}
pairs(data, main = "Scatterplot", col = "blue")
```

**Answer:**

- rank, discipline, sex

> These categorical vars are not well represented in a scatterplot. 

- yrs.since.phd vs salary

> Clear positive correlation, higher salary with more years since phd.

- rank vs salary

> Slightly linear association indicating higher rank correlates with higher salary.

- sex vs salary

> ANOVA must be used to determine if there is significant difference in salaries between genders.

---

### Model Fitting

4. Fit a multiple linear regression model (MLR) with \texttt{yrs.since.phd}, \texttt{discipline}, \texttt{rank}, and \texttt{sex} as predictors. Write down the fitted regression equations for each category (e.g., Female, Assistant Professor, theoretical departments). There are 12 categories in total

**Code:**

```{r}
model <- lm(salary ~ yrs.since.phd + discipline + rank + sex, data = Salaries)
summary(model)
```

**Answer:**

- Fitted equation for salary

$\hat{salary} = 67884.32 + 61.01 * (yrs.since.phd) + 13937.47*(disciplineB) + 13104.15*(AssocProf) + 46032.55*(Prof) + 4349.37*(Male)$

1. \textbf{Female, Assistant, Theoretical}

$\hat{salary} = 67884.32 + 61.01 \times (yrs.since.phd)$

2. \textbf{Female, Assistant, Applied}

$\hat{salary} = 67884.32 + 61.01 \times (yrs.since.phd) + 13937.47 \times (disciplineB)$

3. \textbf{Female, Associate, Theoretical}

$\hat{salary} = 67884.32 + 61.01 \times (yrs.since.phd) + 13104.15 \times (AssocProf)$

4. \textbf{Female, Associate, Applied}

$\hat{salary} = 67884.32 + 61.01 \times (yrs.since.phd) + 13104.15 \times (AssocProf) + 13937.47 \times (disciplineB)$

5. \textbf{Female, Full, Theoretical}

$\hat{salary} = 67884.32 + 61.01 \times (yrs.since.phd) + 46032.55 \times (Prof)$

6. \textbf{Female, Full, Applied}

$\hat{salary} = 67884.32 + 61.01 \times (yrs.since.phd) + 46032.55 \times (Prof) + 13937.47 \times (disciplineB)$

7. \textbf{Male, Assistant, Theoretical}

$\hat{salary} = 67884.32 + 61.01 \times (yrs.since.phd) + 4349.37 \times (Male)$

8. \textbf{Male, Assistant, Applied}

$\hat{salary} = 67884.32 + 61.01 \times (yrs.since.phd) + 13937.47 \times (disciplineB) + 4349.37 \times (Male)$

9. \textbf{Male, Associate, Theoretical}

$\hat{salary} = 67884.32 + 61.01 \times (yrs.since.phd) + 13104.15 \times (AssocProf) + 4349.37 \times (Male)$

10. \textbf{Male, Associate, Applied}

$\hat{salary} = 67884.32 + 61.01 \times (yrs.since.phd) + 13104.15 \times (AssocProf) + 13937.47 \times (disciplineB) + 4349.37 \times (Male)$

11. \textbf{Male, Full, Theoretical}

$\hat{salary} = 67884.32 + 61.01 \times (yrs.since.phd) + 46032.55 \times (Prof) + 4349.37 \times (Male)$

12. \textbf{Male, Full, Applied}

$\hat{salary} = 67884.32 + 61.01 \times (yrs.since.phd) + 46032.55 \times (Prof) + 13937.47 \times (disciplineB) + 4349.37 \times (Male)$

---

5. State the model assumptions in the previous regression model.

**Answer:**

- Linear

> Dependent var (salary) and independent vars (remaining 4 vars) should have linear relationship.

- Independence

> Residuals assumed to be independent, obervations should not affect each other.

- Homoscedasticity

> Varience of residuals must be constant across independent vars

- Normality of Residuals

> Residuals should have a normal distributions.

- No Multicolinearity

> Independent vars should not have high correlation with each other.

---

6. Now fit another MLR with \texttt{yrs.since.phd}, \texttt{discipline}, \texttt{sex} and their interactions. Write down the fitted regression equations for each category

**Code:**

```{r}
interact_model <- lm(salary ~ yrs.since.phd * discipline * sex, data = Salaries)
summary(interact_model)
```

**Answer:**

1. \textbf{Female, Theoretical}

$\hat{salary} = 66195.4 + 1306.8 \times (yrs.since.phd)$

2. \textbf{Female, Applied}

$\hat{salary} = 66195.4 + 1306.8 \times (yrs.since.phd) + 2677.9 + 1397.1 \times (yrs.since.phd)$

3. \textbf{Male, Theoretical}

$\hat{salary} = 66195.4 + 1306.8 \times (yrs.since.phd) + 22688.4 − 475.8 \times (yrs.since.phd)$

4. \textbf{Male, Applied}

$\hat{salary} = 66195.4 + 1306.8 \times (yrs.since.phd) + 2677.9 + 22688.4 + 1397.1 \times (yrs.since.phd) - 475.8 (yrs.since.phd) + 2194$

---

## Non-linear Regression: An Simulated Example

Suppose the response $y$ depends on the predictor $t$ in the following form:
$$y = \alpha \exp(-\beta t) + \epsilon,$$
where $\epsilon \sim \mathrm{N}(0, \sigma^2)$, and the true $\alpha$, $\beta$, and $\sigma^2$
are 15, 2 and 0.16, respectively. First, let's simulate some data points from this nonlinear model:

**Code:**

```{r}
alpha = 15; beta = 2; sigma.sq = 0.09
n <- 50
t <- seq(0, 2, len = 50)
set.seed(123)
y <- alpha * exp(-beta * t) + rnorm(n, sd = sqrt(sigma.sq))
data <- data.frame(y = y, t = t)

plot(t, y, las = 1, pch = 16)
lines(t, alpha * exp(-beta * t), type = "l", col = "blue")
legend("topright", legend = c("Data", "True"), pch = c(16, NA), lty = c(NA, 1),
       col = c("black", "blue"), bty = "n")
```

7. Use the \texttt{nls} function to obtain nonlinear least-squares estimates $\hat{\alpha}$, $\hat{\beta}$, and $\hat{\sigma}^2$. To use \texttt{nls}, provide formula = y $\sim$ alpha * exp(-beta * t), start = list(alpha = alpha_0, beta = beta_0), where alpha_0 and  beta_0 are initial guesses of the parameters $\alpha$ and $\beta$


**Code:**

```{r}
alpha_0 <- 10
beta_0 <- 1

nonlinear_fit <- nls(y ~ alpha * exp(-beta * t), data = data, 
                     start = list(alpha = alpha_0, beta = beta_0))
summary(nonlinear_fit)

alpha_hat <- coef(nonlinear_fit)['alpha']
beta_hat <- coef(nonlinear_fit)['beta']

# residual varience
res <- residuals(nonlinear_fit)
sigma_hat_sq <- var(res)

# estimated vals
print(c("Estimated alpha:", alpha_hat))
print(c("Estimated beta:", beta_hat))
print(c("Estimated sigma squared:", sigma_hat_sq))
```

```{r}
# plotting estimated curve
plot(t, y, las = 1, pch = 16, main = "Nonlinear Least Squares Fit")
lines(t, alpha_hat * exp(-beta_hat * t), col = "orange", lwd = 2)
lines(t, 15 * exp(-2 * t), col = "blue", lwd = 2, lty = 2) # to compare
legend("topright", legend = c("Data", "Estimated", "True"), 
       pch = c(16, NA, NA), lty = c(NA, 1, 2), 
       col = c("black", "orange", "blue"), bty = "n")
```

**Answer:**

Estimated values:

$\hat{\alpha} = 15.0962242708225, \hat{\beta} = 2.0113176, \hat{\sigma}^2 = 0.0765514384$

Estimate equation:

$\hat{y} = 15.0962 e^{-2.011376t} + \epsilon, \quad \epsilon \sim N(0, 0.07655)$

---

8. Write down the fitted equation and the estimated variance $\hat{\sigma}^2$  

**Answer:**

Fitted Equation
$\hat{y} = 15.0962 e^{-2.011376t} + \epsilon, \quad \epsilon \sim N(0, 0.07655)$

Estimated Variance
$\hat{\sigma}^2 = 0.07655$

---

9. Apply the natural log transformation to the simulated response, then fit a simple linear regression. Back-transform to obtain the fit on the original scale

**Code:**

```{r}
# simple linear reg
linear_fit <- lm(log(y) ~ t, data = data)
summary(linear_fit)

# coeff
alpha_log <- exp(coef(linear_fit)[1])
beta_log <- -coef(linear_fit)[2]

# back transform
plot(t, y, pch = 16, main = "Log-Transformed Linear Fit (Back-Transformed)")
lines(t, alpha_log * exp(-beta_log * t), col = "red", lwd = 2)
```

**Answer:**

- Nonlinear Model

$ln(y) = ln(\alpha) - \beta t + \epsilon', \epsilon'  \sim N(0, \sigma'^2)$

---

10. Comparing the nonlinear regression method and the linear regression with log-transformed response, which method would you prefer in this example? Explain your answer

**Answer:**

- NLS as it can give parameter estimates with greater accuracy without being affected by transformations.

---
