---
title: 'DSA 8020 R Lab 3: Multiple Linear Regression II'
author: "saransh rakshak"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Percentage of Body Fat and Body Measurements

Age, weight, height, and 10 body circumference measurements are recorded for 252 men. Each man's percentage of body fat was accurately estimated by an underwater weighing technique.


*Data Source:* Johnson R. *Journal of Statistics Education* v.4, n.1 (1996)


## Load the dataset

**Code:**

```{r}
library(faraway)
data(fat)
head(fat)
```

For the purposes of this lab, we will use only the following variables for conducting data analysis:

1. $y$ \texttt{brozek}: Percent body fat using Brozek's equation
$$\frac{457}{\text{Density}} - 414.2$$

2. $x_{1}$ \texttt{age}: Age (yrs); 

3. $x_2$\texttt{weight}: Height (inches);

4. $x_3$ \texttt{height}: Height (inches);

5. $x_4$\texttt{chest}: Chest circumference (cm);

6. $x_5$ \texttt{abdom}: Abdomen circumference (cm) at the umbilicus and level with the iliac crest

**Code:**

You can use the code below to extract these variables:

```{r}
vars <- c("brozek", "age", "weight", "height", "chest", "abdom")
data <- fat[, vars]
```

---
 
## Exploratory Data Analysis

### Numerical summary

1. Use \texttt{summary} commend to produce various numerical summaries of each of the 6 variables under consideration  

**Code:**

```{r}
summary(data)
```

---

### Graphical summary

2. Make a boxplot for each variable 

**Code:**

```{r}
par(mfrow = c(2, 3))
for (var in vars) {
  boxplot(data[[var]], main = var, col = "lightblue")
}
```

---

3. Briefly discuss the shape of the distribution of each variable

**Answer:**

Brozek 

- Slight right skew.
- Few high outliers on upper end.

Age

- Distribution appears symetrical.

Weight

- Right-skewed distribution.
- Few high-value outliers.

Height

- Approximately symmetric.
- Single outlier on lower end.

Chest

- Right skew.
- Few high value outliers.

Abdomen

- Right skew.
- Multiple high value outliers.

---

4. Create a scatterplot matrix to explore the inter-dependence between these variables 

**Code:**

```{r}
pairs(data, main = "Scatterplot", col = "blue")
```

---

## General Linear F-Test

Suppose a researcher would like to compare the "Full" model using all the 5 predictors and a "reduce" model where only $x_{1}$ (\texttt{age}) and $x_{5}$ (\texttt{abdom}) are used by performing a general linear F-test:

5. Write down the null and the alternative hypotheses

**Answer:**

Null Hypothesis ($H_{0}$)

> The additional predictors do not improve the model, thus reduced model fits the data as well as the full.

Alt Hypothesis ($H_{1}$)

> The full model greatly fits the data better than the reduced model, showing that the additional predictors are vital.

---

6. Fit the full model and write down the fitted linear regression equation.

**Code:**

```{r}
full_model <- lm(brozek ~ age + weight + height + chest + abdom, data = data)
summary(full_model)
```

**Answer:**

$\hat{brozek} = \beta_0 + \beta_1 * age + \beta_2 * weight + \beta_3 * height + \beta_4 * chest + \beta_5 * abdomen$

---

7. Fit the reduced model and write down the fitted linear regression equation.

**Code:**

```{r}
reduced_model <- lm(brozek ~ age + abdom, data = data)
summary(reduced_model)
```

**Answer:**

$\hat{brozek} = \beta_0 + \beta_1 * age + \beta_2 * abdomen$

---


8. Perform a general linear F-test and state the conclusion at $\alpha = 0.05$

**Code:**

```{r}
anova(reduced_model, full_model)
```

**Answer:**

We reject null hypothesis as p-value of 1.543e-08 much lower than our significance level alpha = 0.05.


---

## Prediction

9. Predict a future response for an individual with \texttt{age} $=54$, \texttt{weight} $=197$, \texttt{height} $=72.25$, \texttt{chest} $=105.375$, and \texttt{abdom} $= 99.325$. Construct a 95% prediction interval.

**Code:**

```{r}
new_data <- data.frame(age = 54, weight = 197, height = 72.25, chest = 105.375, abdom = 99.325)
predict(full_model, newdata = new_data, interval = "prediction", level = 0.95)
```


**Answer:**

For a patient with these characteristics, we can expect a body fat percentage of 22.4%.

---

10. Construct a 95% confidence interval for the mean response of percent body fat with \texttt{age} $=54$, \texttt{weight} $=197$, \texttt{height} $=72.25$, \texttt{chest} $=105.375$, and \texttt{abdom} $= 99.325$.
 
 **Code:**

```{r}
predict(full_model, newdata = new_data, interval = "confidence", level = 0.95)
```
 
---

## Multicollinearity


11. Make the scatterplot matrix and compute the correlation matrix for all 6 variables (including the response).

**Code:**

```{r}
pairs(data, main = "Scatterplot Matrix")
cor(data)
```

---

12. Calculate VIF and briefly discuss your finding

**Code:**

```{r}
vif(full_model)
```

**Answer:**

No Multicollinearity:

- Age (1.42)
- Height (1.37)

Moderate Multicollinearity:

- Chest (7.98)

Severe Multicollinearity:

- Abdoment (9.38)
- Weight (10.05)




