---
title: 'DSA 8020 R Lab 4: Model Selection and Model Checking'
author: "Whitney"
output:
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Savings rates in 50 countries

The savings data frame has 50 rows (countries) and 5 columns (variables):

1. \texttt{sr}: savings rate - personal saving divided by disposable income
*This variable will be used as the response*
2. \texttt{pop15}: percent population under age of 15
3. \texttt{pop75}: percent population over age of 75
4. \texttt{dpi}: per-capita disposable income in dollars
5. \texttt{ddpi}: percent growth rate of dpi

The data is averaged over the period 1960-1970. 



*Data Source:* Belsley, D., Kuh. E. and Welsch, R. (1980) *Regression Diagnostics* Wiley.


Load the dataset

**Code:**

```{r}
data(savings, package = "faraway")
head(savings)
```

### 1. Perform the best subset selection and select the "best" model using $R^{2}_{adj}$

**Code:**

```{r,message=FALSE}
library(tidyverse)
library(caret)
library(leaps)
models <- regsubsets(sr ~ ., data = savings)
(res.sum <- summary(models))
criteria <- data.frame(
Adj.R2 = res.sum$adjr2,
Cp = res.sum$cp,
BIC = res.sum$bic)
criteria
```

**Answer:**

We would select the model that includes \texttt{pop15}, \texttt{pop75}, and \texttt{ddpi} based on $R^{2}_{adj}$. 


### 2. Perform a stepwise selection using *AIC*

**Code:**

```{r}
full <- lm(sr~., data = savings)
step(full, direction = "both")
```

**Answer:**

We would again select the model that includes \texttt{pop15}, \texttt{pop75}, and \texttt{ddpi}.

### 3. Perform a general linear $F$-test (with $\alpha = 0.1$) to choose between the full model (i.e., using the all 4 predictors) and the reduce model that include \texttt{pop15}, \texttt{pop75}, and \texttt{ddpi} as the predictors 

**Code:**

```{r}
reduce <- lm(sr ~ pop15 + pop75 + ddpi, data = savings)
anova(reduce, full)
```

**Answer:**

The p-value is greater than $\alpha$, therefore we fail to reject $H_{0}$, meaning that we do not have sufficient evidence to include \texttt{dpi}.

### 4. Make a residual plot of the model selected by *AIC* and comment the model assumptions

**Code:**

```{r}
aicModel <- step(full, direction = "both", trace = F)
plot(aicModel, which = c(1, 3))
```

**Answer:**

The linearity assumption and constant variance assumption are reasonable based on the residual plots.

### 5. Use both histogram and qqplot to examine the normality assumption on error 

**Code:**

```{r}
par(mfrow = c(1, 2))
hist(aicModel$residuals, 10, las = 1, col = "lightblue")
qqnorm(aicModel$residuals, pch = 16, cex = 0.75, col = "blue")
qqline(aicModel$residuals)
```

**Answer:**

The normality assumption on error term is reasonable.

### 6. Calculate the leverage values to check if there is any high leverage points (i.e., $h>\frac{2p}{n}$)

**Code:**

```{r}
lev <- hatvalues(aicModel)
p <- length(coef(aicModel)); n <- dim(savings)[1]
which(lev >= (2 * p) / n)
row.names(savings)[which(lev >= (2 * p) / n)]
```

**Answer:**

There are three countries with high level points `r row.names(savings)[which(lev >= (2 * p) / n)]`

### 7. Compute jackknife residuals to identify outlier(s)

**Code:**

```{r}
jack <- rstudent(aicModel)

par(las = 1)
plot(jack, pch = 16, cex = 0.8, col = "blue", main =" Jacknife Residuals ",
     xlab = "", ylab = "")
abline(h = 0, lty = 2, col = "gray")
```

**Answer:**

There is no obvious outlier

### 8. Identifying influential observations by computing DFFITS

**Code:**

```{r,message=FALSE}
library(olsrr)
ols_plot_dffits(aicModel)
row.names(savings)[c(23, 46, 49)]
```

**Answer:**

`r row.names(savings)[c(23, 46, 49)]`



